---
output:
  github_document:
    html_preview: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# emptyviz

<!-- badges: start -->
[![R-CMD-check](https://github.com/mkthalmann/emptyviz/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mkthalmann/emptyviz/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

A `ggplot2` theme (`theme_mt()`) plus a set of composable geoms and plot
builders used across the author's experimental projects: mean &plusmn; 1 SD
violin/half-violin/split-violin geoms for raw observations, and
`ggdist`-based halfeye/ridge/coefficient-grid, location-scale, and log Bayes
Factor forest plot builders for posterior draws from Bayesian models
(typically fit with `brms`).

This package is not intended for general audiences - it's published so that
the specific analyses that depend on it are reproducible by anyone who
installs the tagged version they were built with.

## Installation

```r
# install.packages("pak")
pak::pak("mkthalmann/emptyviz")
```

To pin the exact version a given analysis was built with (recommended for
reproducing a specific plot), install a tag or commit instead of the
default branch:

```r
pak::pak("mkthalmann/emptyviz@v0.1.0")
```

`theme_mt()` defaults to the `"Roboto Condensed"` font family, which this
package does not install - it must already be available on the system for
text to render as intended. Most graphics devices fall back to a different
font silently rather than erroring, so a missing font won't be obvious;
install it, or pass `theme_mt(base_family = "")`/another available family,
if reproducing a plot's exact appearance matters.

## Usage

```{r example, message = FALSE}
library(emptyviz)
library(ggplot2)

# sets theme_mt() as the active ggplot2 theme
use_theme_mt()
```

```{r violin-example, fig.width = 7, fig.height = 5}
set.seed(42)
demo_data <- rbind(
  data.frame(group = "A", value = rnorm(80, 0, 1)),
  data.frame(group = "B", value = rnorm(80, 1.5, 0.6)),
  data.frame(group = "C", value = rnorm(80, -0.5, 1.8))
)

ggplot(demo_data, aes(x = group, y = value, fill = group)) +
  geom_violin_sd() +
  labs(
    title = "geom_violin_sd()",
    subtitle = "aura (full density) + mean &plusmn;1 SD band",
    x = NULL, y = NULL
  ) +
  guides(fill = "none")
```

```{r bf-example, fig.width = 7, fig.height = 4}
bf_data <- data.frame(
  contrast = c("A vs. B", "C vs. D", "E vs. F"),
  log_bf = c(8.2, 1.1, -0.4)
)

plot_bf_forest(
  bf_data,
  contrast = contrast,
  log_bf = log_bf,
  direction_labels = c("supports difference", "supports equivalence")
)
```

See `vignette("geoms-and-theme")` for the theme and the three `_sd` violin
geoms, and `vignette("bayesian-plots")` for the halfeye/ridge/coefficient-grid,
location-scale, and Bayes Factor forest plot builders.

## Dark mode

`theme_mt(dark = TRUE)` is a dark-background variant - transparent
plot/panel background, light text/gridline colors, and `dark_mt_colors5` in
place of `mt_colors5` as the discrete palette - for a plot rendered directly
against a dark page. `theme_mt()`'s regular (`dark = FALSE`) output is
unaffected by this argument existing.

In a Quarto document, `use_theme_mt()` also registers automatic dual
light/dark figure rendering: set the `dual_render` chunk option to `TRUE`
(per chunk, or as a project-wide `knitr: opts_chunk: dual_render: true`
default) and every `ggplot`/`patchwork` figure in that chunk renders twice -
once normally, once with the dark variant - wrapped for Quarto's
`.light-content`/`.dark-content` toggle, so a light/dark theme-switching site
shows the right one without any client-side re-rendering. Chunks that don't
set the option are unaffected.

## Development

```r
devtools::load_all()
devtools::test()
devtools::check()
```
