---
title: "Plotting Bayesian posteriors: halfeye, location-scale, and BF forest plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting Bayesian posteriors: halfeye, location-scale, and BF forest plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6
)
```

```{r setup, message = FALSE}
library(emptyviz)
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
library(patchwork)
use_theme_mt()
```

## The model behind `believe_projection_draws`

This vignette's data comes from the distributional `brms` model
Thalmann/Matticchio fit to `believe_projection`
(see `vignette("geoms-and-theme")` for the same data's raw-observation
plots). Both the judgment mean *and* its spread are modeled as a function of
condition — the kind of model this package's Bayesian plot builders exist
for:

```{r real-model, eval = FALSE}
# not run here - 8 chains x 10,000 iterations is well beyond what a vignette
# should fit live. The draws used below were extracted from this exact
# cached model by data-raw/believe-projection-posteriors.R; see
# ?believe_projection_draws et al. for how.
mod <- brm(
  bf(
    judgment ~ negation * scenario * trigger +
      (1 + scenario * negation * trigger | id) +
      (1 + scenario * negation | item),
    sigma ~ negation * scenario + trigger
  ),
  data = d, # the cleaned believe_projection from vignette("geoms-and-theme")
  family = gaussian(),
  prior = c(
    prior(normal(0, .5), class = Intercept, lb = -2, ub = 2),
    prior(normal(0, .5), class = Intercept, dpar = sigma),
    prior(normal(0, 1), class = b, lb = -3, ub = 3),
    prior(normal(0, .25), class = b, dpar = sigma),
    prior(normal(0, .5), class = sd)
  ),
  iter = 10000,
  chains = 8
)
```

## `plot_ridge_hdi()` / `plot_coef_grid_hdi()`

Two layout wrappers around one shared layer builder, `layer_halfeye_hdi()`
(a `ggdist::stat_slab()` + `ggdist::stat_pointinterval()` "halfeye" block:
an HDI-shaded density plus a multi-width ETI point-interval per category).
Both expect long-format posterior draws — one row per draw per category, as
produced by `tidybayes::gather_emmeans_draws()` or `spread_draws()` — not
raw observations, unlike the `_sd` violin geoms in `vignette("geoms-and-theme")`.

- `plot_ridge_hdi()`: every category stacked as a row in one shared-scale
  panel, for when the categories themselves are what's being compared.
- `plot_coef_grid_hdi()`: each category gets its own free-scale panel in a
  grid, for coefficient posteriors that differ wildly in natural magnitude.

`believe_projection_draws` (see `?believe_projection_draws`) holds exactly
this shape already: long-format marginal posterior draws for
`scenario`\*`negation`\*`trigger`, for both the `mu` (judgment) and `sigma`
(spread) submodels, extracted via `emmeans() |> gather_emmeans_draws()`.

```{r ridge-data}
ridge_draws <- believe_projection_draws |>
  filter(dpar == "mu") |>
  mutate(cond = paste0(scenario, "-", negation))
```

### `plot_ridge_hdi()`

```{r ridge-basic, fig.height = 7}
plot_ridge_hdi(
  ridge_draws,
  category = cond,
  facet = trigger,
  facet_nrow = 2,
  value_limits = c(-2.3, 2.3),
  value_breaks = c(-2, -1, 0, 1, 2)
)
```

### `category_reorder`

Default `TRUE` sorts rows by mean value (descending); `FALSE` keeps
`category`'s existing factor-level order instead — useful when the order
itself is meaningful (e.g. matching a fixed condition sequence elsewhere in
a paper).

```{r ridge-reorder, fig.height = 5}
p_reorder_on <- plot_ridge_hdi(
  ridge_draws |> filter(trigger == "*again*"),
  category = cond,
  value_limits = c(-2.3, 2.3)
) +
  labs(x = "Reordered (default)")

p_reorder_off <- plot_ridge_hdi(
  ridge_draws |> filter(trigger == "*again*"),
  category = cond,
  category_reorder = FALSE,
  value_limits = c(-2.3, 2.3)
) +
  labs(x = "Original factor order")

p_reorder_on + p_reorder_off
```

### `value_transform`

The `sigma` submodel is fit on the log scale — pass `value_transform = exp`
to back-transform before plotting, matching an `aes(y = exp(.value))`-style
back-transformation done by hand. `believe_projection_draws`' `dpar ==
"sigma"` rows are exactly that submodel's marginal draws.

```{r ridge-transform, fig.height = 5}
sigma_draws <- believe_projection_draws |>
  filter(dpar == "sigma") |>
  mutate(cond = paste0(scenario, "-", negation))

plot_ridge_hdi(
  sigma_draws |> filter(trigger == "*again*"),
  category = cond,
  value_transform = exp,
  ylab = "Posterior marginal SD ±HDI<sub>95</sub> ±ETI<sub>50;90;95</sub>"
)
```

### `plot_coef_grid_hdi()`

Free-scale small multiples for a set of model coefficients — the real
sum-coded fixed effects from both submodels (`believe_projection_coef_draws`,
see `?believe_projection_coef_draws`) differ wildly enough in natural
magnitude that a shared axis would flatten the small ones, exactly the case
this layout is for.

```{r coef-grid-data}
mu_coef_draws <- believe_projection_coef_draws |> filter(dpar == "mu")
sigma_coef_draws <- believe_projection_coef_draws |> filter(dpar == "sigma")
```

```{r coef-grid-basic, fig.height = 8}
plot_coef_grid_hdi(mu_coef_draws, category = coef, scale = 1.4)
```

The `sigma` submodel's coefficients (fewer terms - no three-way interaction
in `sigma ~ negation*scenario + trigger`):

```{r coef-grid-sigma, fig.height = 5}
plot_coef_grid_hdi(sigma_coef_draws, category = coef, scale = 1.4, ncol = 3)
```

`hline` defaults to `0` (a common null-effect reference line for coefficient
grids); pass a different value or `NULL` to move or omit it.

```{r coef-grid-no-hline, fig.height = 6}
plot_coef_grid_hdi(mu_coef_draws, category = coef, scale = 1.4, hline = NULL)
```

## `plot_location_scale()`

A joint (posterior marginal mean) x (posterior marginal standard deviation)
plot: one point + a shaded bivariate credible ellipse per condition, plus an
optional sigma_max ceiling-effect reference curve — `y = sqrt((x -
lower)(upper - x))`, the maximum SD a bounded response scale permits at a
given mean. Lets a reader check whether a condition's spread is close to or
far from that ceiling directly, rather than computing it by hand for one
condition at a time.

No custom `Stat` here either — the bivariate region is
`ggplot2::stat_ellipse()`, already in ggplot2 core.

### Why the draws have to be paired

`data` must contain the location and scale draws already **paired** - one
row per (category[, shape], `.draw`), where the location value and the
scale value on that row came from the *same* posterior iteration. That's
different from independently summarizing the two submodels (e.g. taking
`mean_hdi()` of each separately) - a genuine join is what lets the ellipse
reflect their *joint* uncertainty (including any correlation between a
condition's location and scale draws) rather than two unrelated marginal
intervals mashed into a fake cross.

In a real analysis, this means calling `tidybayes::gather_emmeans_draws()`
twice - once on the location submodel, once with `dpar = "sigma"` - and
joining the results on `.draw` (plus whatever grouping columns identify a
condition):

```{r location-scale-real-analysis, eval = FALSE}
library(brms)
library(emmeans)
library(tidybayes)

mod <- brm(
  bf(rating ~ condition, sigma ~ condition),
  family = gaussian(),
  data = my_data
)

# one row per (condition, .draw)
loc_draws <- mod |>
  emmeans(~condition) |>
  gather_emmeans_draws()

# dpar = "sigma" pulls the scale submodel instead - fit (and so returned)
# on the log scale, hence sigma_transform = exp below
sig_draws <- mod |>
  emmeans(~condition, dpar = "sigma") |>
  gather_emmeans_draws()

# the join is what pairs each location draw with the scale draw from the
# *same* posterior iteration, rather than treating them as independent
paired_draws <- inner_join(
  loc_draws, sig_draws,
  by = c("condition", ".draw"),
  suffix = c("_location", "_sigma")
)

plot_location_scale(
  paired_draws,
  category = condition,
  location = .value_location,
  sigma = .value_sigma,
  sigma_transform = exp,
  bounds = c(1, 7) # e.g. a 1-7 Likert-type response scale
)
```

`believe_projection_draws` already has both submodels' draws stacked long
(a `dpar` column instead of two separate objects) - pivoting `dpar` wide
recovers exactly the `paired_draws` shape above. This only works because
`data-raw/believe-projection-posteriors.R` thinned `mu`/`sigma` down to the
*same* shared set of `.draw` indices rather than independently subsampling
each - the pairing `plot_location_scale()` depends on is a property of
which MCMC iteration a draw came from, so keeping mismatched indices per
submodel would silently produce a fake, unpaired cross.

```{r location-scale-data}
loc_scale_draws <- believe_projection_draws |>
  pivot_wider(names_from = dpar, values_from = .value)
```

`ellipse_level` defaults to `c(.5, .95)` - a narrow "core" credible ellipse
nested inside a wider outer one, told apart by alpha alone (darker/more
opaque core, fainter outer band), the same multi-width idea
`layer_halfeye_hdi()` uses for its point-intervals. Pass a single level
(e.g. `ellipse_level = .95`) to go back to one ellipse per condition.

```{r location-scale-basic, fig.height = 6}
plot_location_scale(
  loc_scale_draws,
  category = scenario,
  location = mu,
  sigma = sigma,
  shape = negation,
  facet = trigger,
  bounds = c(-2, 2)
)
```

## `plot_bf_forest()` / `prepare_bf_contrasts()` / `layer_bf_evidence_scale()`

A log Bayes Factor forest plot: one row per contrast, point &plusmn;
optional SE, colored by sign, with a shaded "weak evidence" region
(`layer_bf_evidence_scale()`, bundled in by default) at the Jeffreys/Lee
`&plusmn;log(3)` "weak/anecdotal evidence" convention. `layer_bf_evidence_scale()`
itself is composable (like `layer_halfeye_hdi()`) - usable on its own on any
plot with a log-BF-valued axis, in either orientation.

### `pairs`: selecting contrasts without hand-parsing strings

`bayestestR::bayesfactor_rope()` computes *every* pairwise contrast in a
reference grid and names each one with an auto-generated `"<left> - <right>"`
string. Picking out just the pairs you actually want to show, and giving
them clean labels, is exactly what's error-prone about doing this by hand
(position-counting words out of the string, then a long hardcoded
`filter(left == "x" & right == "y" | ...)` chain). `plot_bf_forest()`'s
`pairs` argument takes that off your plate: give it a plain list of
`c(left, right)` pairs, in your own words, and it filters + labels +
reorders for you - and errors, naming the exact pair, if one isn't found,
instead of the filter silently matching nothing. You don't need to know
which order `bayesfactor_rope()`'s reference grid happened to enumerate a
given pair in either - a pair matched in the opposite order gets relabeled
*and* sign-flipped to the order you asked for, automatically.

`believe_projection_bf` (see `?believe_projection_bf`) is real
`bayestestR::bayesfactor_rope()` output — one row per pairwise contrast
among the 8 `scenario`\*`negation` cells (`trigger` marginalized out, hence
each raw `contrast` string's trailing `" NA"`), for both submodels. There's
no SE column at all, which is the normal case for this kind of Bayes Factor
(unlike a repeated bridge-sampling estimate) and needs nothing special to
work. `bf_pairs` below picks out the 9 contrasts of interest among the 8
cells' pairwise combinations, as plain `c(left, right)` pairs.

```{r bf-forest-data}
mu_bf <- believe_projection_bf |> filter(dpar == "mu")

bf_pairs <- list(
  c("with undef", "without undef"),
  c("with undef", "with critical"),
  c("with true", "without true"),
  c("with false", "without false"),
  c("with critical", "without critical"),
  c("with false", "with critical"),
  c("without false", "without critical"),
  c("without true", "with false"),
  c("with true", "without false")
)
```

```{r bf-forest-basic, fig.height = 6}
plot_bf_forest(mu_bf, contrast = contrast, log_bf = log_BF, pairs = bf_pairs)
```

The trailing `" NA"` (emmeans' placeholder for a reference-grid variable
marginalized out via e.g. `at = list(trigger = NA)`) is stripped
automatically; see `prepare_bf_contrasts()` - the function `plot_bf_forest()`
calls internally - if you want to inspect the split-and-labeled table
directly, or reuse it outside a plot.

If your data already has one clean, pre-labeled row per contrast (no
`bayestestR`-style string to split), just omit `pairs` and pass that column
straight to `contrast`.

### `direction_labels`

Naming what a positive/negative log-BF supports turns the plot into a
self-contained legend for the sign convention, instead of requiring the
reader to already know which direction means what. The arrows and labels
sit right at the top edge of the panel, out of the way of the data, adding
essentially no extra space to the figure.

```{r bf-forest-arrows, fig.height = 6}
plot_bf_forest(
  mu_bf,
  contrast = contrast,
  log_bf = log_BF,
  pairs = bf_pairs,
  direction_labels = c("supports difference", "supports equivalence")
)
```

### `secondary_axis`

A dual log/linear Bayes Factor axis - opt-in, since a log-only axis is
perfectly fine on its own.

```{r bf-forest-secaxis, fig.height = 6}
plot_bf_forest(
  mu_bf,
  contrast = contrast,
  log_bf = log_BF,
  pairs = bf_pairs,
  direction_labels = c("supports difference", "supports equivalence"),
  secondary_axis = TRUE
)
```

### `layer_bf_evidence_scale(orientation = "y")`

The same annotation bundle on a continuous-x sweep (e.g. log-BF as a
function of a prior's width), with log-BF on the *y*-axis instead - the
rect, arrows, and labels all rotate to hug the right edge of the panel
instead of the top.

```{r bf-evidence-scale-y, fig.height = 5}
sweep_data <- data.frame(
  prior_sd = seq(0.1, 1.5, by = 0.2),
  log_bf = c(6.2, 5.1, 3.8, 2.5, 1.4, 0.3, -0.6, -1.5)
)

ggplot(sweep_data, aes(x = prior_sd, y = log_bf)) +
  layer_bf_evidence_scale(
    orientation = "y",
    range = c(-3, 8),
    direction_labels = c("supports difference", "supports equivalence")
  ) +
  geom_line(color = mt_colors[1]) +
  geom_point(color = mt_colors[1], size = 3) +
  labs(x = "Prior standard deviation", y = "log Bayes factor") +
  # unlike orientation = "x"'s top edge, the default right-side expansion
  # isn't quite wide enough for orientation = "y"'s rotated label + arrow -
  # this is the caller's own coord/margin choice, not something
  # layer_bf_evidence_scale() should impose on every plot itself.
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(5.5, 55, 5.5, 5.5, "pt"))
```
