---
title: "Plotting Bayesian posteriors: halfeye, location-scale, and BF forest plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting Bayesian posteriors: halfeye, location-scale, and BF forest plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6
)
```

```{r setup, message = FALSE}
library(emptyviz)
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
library(patchwork)
use_theme_mt()
```

## `plot_ridge_hdi()` / `plot_coef_grid_hdi()`

Two layout wrappers around one shared layer builder, `layer_halfeye_hdi()`
(a `ggdist::stat_slab()` + `ggdist::stat_pointinterval()` "halfeye" block:
an HDI-shaded density plus a multi-width ETI point-interval per category).
Both expect long-format posterior draws — one row per draw per category, as
produced by `tidybayes::gather_emmeans_draws()` or `spread_draws()` — not
raw observations, unlike the `_sd` violin geoms in `vignette("geoms-and-theme")`.

- `plot_ridge_hdi()`: every category stacked as a row in one shared-scale
  panel, for when the categories themselves are what's being compared.
- `plot_coef_grid_hdi()`: each category gets its own free-scale panel in a
  grid, for coefficient posteriors that differ wildly in natural magnitude.

Synthetic data below stands in for a Bayesian distributional model's
posterior: four scenarios crossed with a 2-level factor, faceted by a
between-item factor, with the middle two scenarios deliberately given wider
spread than the outer two, to illustrate a location+scale pattern.

```{r ridge-data}
set.seed(42)
conditions <- expand_grid(
  scenario = c("true", "false", "undefined", "critical"),
  negation = c("without", "with"),
  trigger = c("again", "stop")
) |>
  mutate(
    mu = case_when(
      scenario == "true" & negation == "without" ~ 1.5,
      scenario == "false" & negation == "with" ~ 1.5,
      scenario == "true" & negation == "with" ~ -1.5,
      scenario == "false" & negation == "without" ~ -1.5,
      TRUE ~ 0
    ),
    # undefined/critical get wider posteriors - the location+scale pattern
    # plot_ridge_hdi() (paired with a second call on the sigma submodel) is
    # meant to make visible
    sigma = if_else(scenario %in% c("undefined", "critical"), .8, .3)
  )

ridge_draws <- conditions |>
  rowwise() |>
  mutate(draws = list(rnorm(600, mu, sigma))) |>
  ungroup() |>
  unnest(draws) |>
  rename(.value = draws) |>
  mutate(cond = paste0(scenario, "-", negation))
```

### `plot_ridge_hdi()`

```{r ridge-basic, fig.height = 7}
plot_ridge_hdi(
  ridge_draws,
  category = cond,
  facet = trigger,
  facet_nrow = 2,
  value_limits = c(-2.05, 2.05),
  value_breaks = c(-2, -1, 0, 1, 2)
)
```

### `category_reorder`

Default `TRUE` sorts rows by mean value (descending); `FALSE` keeps
`category`'s existing factor-level order instead — useful when the order
itself is meaningful (e.g. matching a fixed condition sequence elsewhere in
a paper).

```{r ridge-reorder, fig.height = 5}
p_reorder_on <- plot_ridge_hdi(
  ridge_draws |> filter(trigger == "again"),
  category = cond,
  value_limits = c(-2.05, 2.05)
) +
  labs(x = "Reordered (default)")

p_reorder_off <- plot_ridge_hdi(
  ridge_draws |> filter(trigger == "again"),
  category = cond,
  category_reorder = FALSE,
  value_limits = c(-2.05, 2.05)
) +
  labs(x = "Original factor order")

p_reorder_on + p_reorder_off
```

### `value_transform`

For a scale (sigma) submodel typically fit on the log scale — pass
`value_transform = exp` to back-transform before plotting, matching a
`aes(y = exp(.value))`-style back-transformation done by hand.

```{r ridge-transform, fig.height = 5}
log_sigma_draws <- conditions |>
  distinct(scenario, negation, trigger, sigma) |>
  mutate(log_sigma_mean = log(sigma)) |>
  rowwise() |>
  mutate(draws = list(rnorm(600, log_sigma_mean, .15))) |>
  ungroup() |>
  unnest(draws) |>
  rename(.value = draws) |>
  mutate(cond = paste0(scenario, "-", negation))

plot_ridge_hdi(
  log_sigma_draws |> filter(trigger == "again"),
  category = cond,
  value_transform = exp,
  ylab = "Posterior marginal SD ±HDI<sub>95</sub> ±ETI<sub>50;90;95</sub>"
)
```

### `plot_coef_grid_hdi()`

Free-scale small multiples for a set of model coefficients — deliberately
simulated here with very different magnitudes/spreads, the case a shared
axis would flatten.

```{r coef-grid-data}
coef_specs <- tibble::tribble(
  ~coef, ~mu, ~sd,
  "Intercept", 1.0, .35,
  "Negation", -.4, .25,
  "false", 1.3, .3,
  "undef", .3, .5,
  "critical", .35, .12,
  "stop", .8, .3,
  "Negation:false", -.9, .3,
  "Negation:undef", .8, .55
) |>
  mutate(coef = fct_inorder(coef))

coef_draws <- coef_specs |>
  rowwise() |>
  mutate(draws = list(rnorm(1200, mu, sd))) |>
  ungroup() |>
  unnest(draws) |>
  rename(.value = draws)
```

```{r coef-grid-basic, fig.height = 6}
plot_coef_grid_hdi(coef_draws, category = coef, scale = 1.4)
```

`hline` defaults to `0` (a common null-effect reference line for coefficient
grids); pass a different value or `NULL` to move or omit it.

```{r coef-grid-no-hline, fig.height = 6}
plot_coef_grid_hdi(coef_draws, category = coef, scale = 1.4, hline = NULL)
```

## `plot_location_scale()`

A joint (posterior marginal mean) x (posterior marginal standard deviation)
plot: one point + a shaded bivariate credible ellipse per condition, plus an
optional sigma_max ceiling-effect reference curve — `y = sqrt((x -
lower)(upper - x))`, the maximum SD a bounded response scale permits at a
given mean. Lets a reader check whether a condition's spread is close to or
far from that ceiling directly, rather than computing it by hand for one
condition at a time.

No custom `Stat` here either — the bivariate region is
`ggplot2::stat_ellipse()`, already in ggplot2 core.

### Why the draws have to be paired

`data` must contain the location and scale draws already **paired** - one
row per (category[, shape], `.draw`), where the location value and the
scale value on that row came from the *same* posterior iteration. That's
different from independently summarizing the two submodels (e.g. taking
`mean_hdi()` of each separately) - a genuine join is what lets the ellipse
reflect their *joint* uncertainty (including any correlation between a
condition's location and scale draws) rather than two unrelated marginal
intervals mashed into a fake cross.

In a real analysis, this means calling `tidybayes::gather_emmeans_draws()`
twice - once on the location submodel, once with `dpar = "sigma"` - and
joining the results on `.draw` (plus whatever grouping columns identify a
condition):

```{r location-scale-real-analysis, eval = FALSE}
library(brms)
library(emmeans)
library(tidybayes)

mod <- brm(
  bf(rating ~ condition, sigma ~ condition),
  family = gaussian(),
  data = my_data
)

# one row per (condition, .draw)
loc_draws <- mod |>
  emmeans(~condition) |>
  gather_emmeans_draws()

# dpar = "sigma" pulls the scale submodel instead - fit (and so returned)
# on the log scale, hence sigma_transform = exp below
sig_draws <- mod |>
  emmeans(~condition, dpar = "sigma") |>
  gather_emmeans_draws()

# the join is what pairs each location draw with the scale draw from the
# *same* posterior iteration, rather than treating them as independent
paired_draws <- inner_join(
  loc_draws, sig_draws,
  by = c("condition", ".draw"),
  suffix = c("_location", "_sigma")
)

plot_location_scale(
  paired_draws,
  category = condition,
  location = .value_location,
  sigma = .value_sigma,
  sigma_transform = exp,
  bounds = c(1, 7) # e.g. a 1-7 Likert-type response scale
)
```

The synthetic draws below stand in for `paired_draws` above - already
paired by construction, since each row is built directly from one `mu`/`sig`
pair per condition, so no join is needed here.

```{r location-scale-data}
set.seed(42)
loc_scale_specs <- expand_grid(
  scenario = c("true", "false", "undefined", "critical"),
  negation = c("without", "with"),
  trigger = c("again", "stop")
) |>
  mutate(
    mu_base = case_when(
      scenario == "true" & negation == "without" ~ 1.5,
      scenario == "false" & negation == "with" ~ 1.5,
      scenario == "true" & negation == "with" ~ -1.5,
      scenario == "false" & negation == "without" ~ -1.5,
      TRUE ~ 0
    ),
    # `trigger` plus a bit of per-condition jitter nudges each condition off
    # its "pure" mu_base - a real dataset practically never has, e.g.,
    # `undefined`/`critical` land on the exact same mean, so a demo where
    # every condition sat exactly on mu_base (and every facet was a near-
    # identical copy of the other) undersold what this plot actually looks
    # like on real data.
    mu = mu_base +
      if_else(trigger == "again", -0.2, 0.3) +
      rnorm(n(), 0, .25),
    # same location+scale pattern as the plot_ridge_hdi() demo above -
    # undefined/critical get wider posteriors than true/false - plus the
    # same idea as `mu`: trigger and per-condition noise vary the *width*
    # too, so conditions don't all carry an identical, interchangeable
    # uncertainty band.
    sig_base = if_else(scenario %in% c("undefined", "critical"), .8, .3),
    sig = pmax(
      .12,
      sig_base +
        if_else(negation == "with", .12, -.05) +
        if_else(trigger == "stop", .08, -.04) +
        rnorm(n(), 0, .12)
    ),
    # the ellipses' own size varies by condition too - not just their
    # center - via the per-row SDs the draws below are sampled from.
    loc_draw_sd = pmax(
      .06,
      if_else(scenario %in% c("undefined", "critical"), .22, .13) + rnorm(n(), 0, .04)
    ),
    sig_draw_sd = pmax(
      .03,
      if_else(scenario %in% c("undefined", "critical"), .11, .06) + rnorm(n(), 0, .02)
    )
  )

loc_scale_draws <- loc_scale_specs |>
  rowwise() |>
  mutate(draws = list(data.frame(
    .draw = 1:500,
    location = rnorm(500, mu, loc_draw_sd),
    sigma = rnorm(500, sig, sig_draw_sd)
  ))) |>
  ungroup() |>
  select(scenario, negation, trigger, draws) |>
  unnest(draws)
```

`ellipse_level` defaults to `c(.5, .95)` - a narrow "core" credible ellipse
nested inside a wider outer one, told apart by alpha alone (darker/more
opaque core, fainter outer band), the same multi-width idea
`layer_halfeye_hdi()` uses for its point-intervals. Pass a single level
(e.g. `ellipse_level = .95`) to go back to one ellipse per condition.

```{r location-scale-basic, fig.height = 6}
plot_location_scale(
  loc_scale_draws,
  category = scenario,
  shape = negation,
  facet = trigger,
  bounds = c(-2, 2)
)
```

`true`/`false` sit tucked well inside their own (narrow) local ceiling near
the scale's edges, and `undefined`/`critical` sit well inside their own
(much higher) local ceiling near the scale's midpoint.

## `plot_bf_forest()` / `prepare_bf_contrasts()` / `layer_bf_evidence_scale()`

A log Bayes Factor forest plot: one row per contrast, point &plusmn;
optional SE, colored by sign, with a shaded "weak evidence" region
(`layer_bf_evidence_scale()`, bundled in by default) at the Jeffreys/Lee
`&plusmn;log(3)` "weak/anecdotal evidence" convention. `layer_bf_evidence_scale()`
itself is composable (like `layer_halfeye_hdi()`) - usable on its own on any
plot with a log-BF-valued axis, in either orientation.

### `pairs`: selecting contrasts without hand-parsing strings

`bayestestR::bayesfactor_rope()` computes *every* pairwise contrast in a
reference grid and names each one with an auto-generated `"<left> - <right>"`
string. Picking out just the pairs you actually want to show, and giving
them clean labels, is exactly what's error-prone about doing this by hand
(position-counting words out of the string, then a long hardcoded
`filter(left == "x" & right == "y" | ...)` chain). `plot_bf_forest()`'s
`pairs` argument takes that off your plate: give it a plain list of
`c(left, right)` pairs, in your own words, and it filters + labels +
reorders for you - and errors, naming the exact pair, if one isn't found,
instead of the filter silently matching nothing. You don't need to know
which order `bayesfactor_rope()`'s reference grid happened to enumerate a
given pair in either - a pair matched in the opposite order gets relabeled
*and* sign-flipped to the order you asked for, automatically.

```{r bf-forest-data}
# stands in for `bayestestR::bayesfactor_rope(pairs(ref_grid), ...)`
# coerced to a data frame - note there's no SE column at all here, which is
# the normal case for this kind of Bayes Factor (unlike a repeated
# bridge-sampling estimate) and needs nothing special to work.
bf_raw <- data.frame(
  contrast = c(
    "true with - true without NA",
    "false with - false without NA",
    "false with - critical with NA",
    "false without - critical without NA",
    "critical with - critical without NA",
    "undef with - undef without NA",
    "true with - false with NA" # not requested below - stays excluded
  ),
  log_BF = c(8.2, 7.9, 5.3, 4.8, 1.1, -0.4, 0.9)
)

bf_pairs <- list(
  c("true with", "true without"),
  c("false with", "false without"),
  c("false with", "critical with"),
  c("false without", "critical without"),
  c("critical with", "critical without"),
  c("undef with", "undef without")
)
```

```{r bf-forest-basic, fig.height = 5}
plot_bf_forest(bf_raw, contrast = contrast, log_bf = log_BF, pairs = bf_pairs)
```

The trailing `" NA"` (emmeans' placeholder for a reference-grid variable
marginalized out via e.g. `at = list(trigger = NA)`) is stripped
automatically; see `prepare_bf_contrasts()` - the function `plot_bf_forest()`
calls internally - if you want to inspect the split-and-labeled table
directly, or reuse it outside a plot.

If your data already has one clean, pre-labeled row per contrast (no
`bayestestR`-style string to split), just omit `pairs` and pass that column
straight to `contrast` - unchanged from before `pairs` existed.

### `direction_labels`

Naming what a positive/negative log-BF supports turns the plot into a
self-contained legend for the sign convention, instead of requiring the
reader to already know which direction means what. The arrows and labels
sit right at the top edge of the panel, out of the way of the data, adding
essentially no extra space to the figure.

```{r bf-forest-arrows, fig.height = 5}
plot_bf_forest(
  bf_raw,
  contrast = contrast,
  log_bf = log_BF,
  pairs = bf_pairs,
  direction_labels = c("supports difference", "supports equivalence")
)
```

### `secondary_axis`

A dual log/linear Bayes Factor axis - opt-in, since a log-only axis is
perfectly fine on its own.

```{r bf-forest-secaxis, fig.height = 5}
plot_bf_forest(
  bf_raw,
  contrast = contrast,
  log_bf = log_BF,
  pairs = bf_pairs,
  direction_labels = c("supports difference", "supports equivalence"),
  secondary_axis = TRUE
)
```

### `layer_bf_evidence_scale(orientation = "y")`

The same annotation bundle on a continuous-x sweep (e.g. log-BF as a
function of a prior's width), with log-BF on the *y*-axis instead - the
rect, arrows, and labels all rotate to hug the right edge of the panel
instead of the top.

```{r bf-evidence-scale-y, fig.height = 5}
sweep_data <- data.frame(
  prior_sd = seq(0.1, 1.5, by = 0.2),
  log_bf = c(6.2, 5.1, 3.8, 2.5, 1.4, 0.3, -0.6, -1.5)
)

ggplot(sweep_data, aes(x = prior_sd, y = log_bf)) +
  layer_bf_evidence_scale(
    orientation = "y",
    range = c(-3, 8),
    direction_labels = c("supports difference", "supports equivalence")
  ) +
  geom_line(color = colors[1]) +
  geom_point(color = colors[1], size = 3) +
  labs(x = "Prior standard deviation", y = "log Bayes factor") +
  # unlike orientation = "x"'s top edge, the default right-side expansion
  # isn't quite wide enough for orientation = "y"'s rotated label + arrow -
  # this is the caller's own coord/margin choice, not something
  # layer_bf_evidence_scale() should impose on every plot itself.
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(5.5, 55, 5.5, 5.5, "pt"))
```
