---
title: "theme_mt() and the SD-band violin geoms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{theme_mt() and the SD-band violin geoms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6
)
```

```{r setup, message = FALSE}
library(emptyviz)
library(ggplot2)
library(dplyr)
library(forcats)
library(gghalves)
library(patchwork)
use_theme_mt()
```

## `believe_projection`

The demos below use `believe_projection`, truth-value judgment data from
Thalmann & Matticchio (2024) — a sentence-picture verification task testing
whether presuppositions triggered by *again*/*stop* under a negated belief
predicate ("Peter isn't certain that Sonja stopped smoking") project
universally or only relative to the attitude holder. `?believe_projection`
has the full column reference; see `vignette("bayesian-plots")` for the
actual distributional model fit to this data.

`sub_exp == "1"` restricts to the main judgment task (dropping training and
a separate world-knowledge norming block); the same three recodes
`believe-projection/scripts/belproj-paper.R` applies before modeling —
rescaling the 0-100 slider to a signed -2/+2 scale, expanding the
abbreviated `"undef"` scenario label, and markdown-italicizing `trigger` so
it renders nicely via `theme_mt()`'s markdown-aware axis text — are exactly
what's reused below and, later, in `vignette("bayesian-plots")`.

```{r believe-projection-clean}
d <- believe_projection |>
  filter(sub_exp == "1") |>
  mutate(
    judgment = (judgment - 50) / 25,
    scenario = gsub("undef", "undefined", scenario),
    scenario = factor(scenario, levels = c("true", "false", "undefined", "critical")),
    negation = factor(negation, levels = c("pos", "neg")),
    negation = fct_recode(negation, without = "pos", with = "neg"),
    trigger = paste0("*", trigger, "*"),
    trigger = factor(trigger, levels = c("*again*", "*stop*"))
  )
```

## `theme_mt()`

Extends `theme_minimal()` with markdown/HTML-aware text (via `ggtext`), a
bottom legend, and an automatic discrete color palette. Same data/labels
below, only the theme differs: `theme_gray()`, `theme_minimal()`, `theme_mt()`.

```{r theme-compare, fig.width = 13, fig.height = 6}
#| fig.alt: >
#|   The same bar chart drawn three times side by side, under theme_gray(),
#|   theme_minimal() and theme_mt(). Each panel shows mean acceptability
#|   ratings for four conditions - Baseline 6.1, Weak Trigger 4.8, Strong
#|   Trigger 2.3, Control 6.5 - with a small standard-error bar and the value
#|   printed above each bar. Only the theme_mt() panel renders the markdown
#|   and HTML entities in its title, subtitle and axis titles; the other two
#|   show the literal markup (&plusmn;, &ndash; and asterisks) as text.
#|   theme_mt() also drops the grey panel background, bolds the title, keeps
#|   only horizontal gridlines, and uses the package palette instead of
#|   ggplot2's default hues.
# theme_mt()'s discrete palette and geom_text() font default are both
# theme-level settings that resolve once, from whichever theme is active
# when the plot's scale/geom defaults are first constructed - not from
# whatever theme ends up attached via `+ theme_X()` afterward. Since this
# vignette's global default is theme_mt() (set by use_theme_mt() above),
# the gray/minimal panels need an explicit scale_fill_hue()/family = "sans"
# to force ggplot2's real defaults instead of silently inheriting
# theme_mt()'s.
theme_compare_data <- data.frame(
  condition = c("Baseline", "Weak Trigger", "Strong Trigger", "Control"),
  mean_rating = c(6.1, 4.8, 2.3, 6.5),
  se = c(0.15, 0.22, 0.28, 0.12)
)
theme_compare_data$condition <- factor(
  theme_compare_data$condition,
  levels = theme_compare_data$condition
)

build_theme_compare_plot <- function() {
  ggplot(theme_compare_data, aes(x = condition, y = mean_rating, fill = condition, color = condition)) +
    geom_col(alpha = .3) +
    geom_errorbar(aes(ymin = mean_rating - se, ymax = mean_rating + se), width = 0.2) +
    labs(
      title = "Acceptability ratings by condition",
      subtitle = "synthetic **data**, *n* = 45 per condition",
      x = "Condition", y = "Mean rating &plusmn;1 SE (1&ndash;7)",
      caption = "Test caption"
    ) +
    guides(fill = "none", color = "none")
}

value_label <- function(...) {
  geom_text(
    aes(label = sprintf("%.1f", mean_rating)),
    vjust = -1.8, color = "black", size = 5,
    ...
  )
}

p_theme_gray <- build_theme_compare_plot() + theme_gray() + scale_fill_hue() + value_label(family = "sans")
p_theme_minimal <- build_theme_compare_plot() + theme_minimal() + scale_fill_hue() + value_label(family = "sans")
p_theme_mt <- build_theme_compare_plot() + theme_mt(base_size = 12) + value_label()

p_theme_gray + p_theme_minimal + p_theme_mt
```

## `geom_violin_sd()` / `geom_half_violin_sd()`

Compound layers built on `geom_violin()`/`gghalves::geom_half_violin()`: a
low-alpha "aura" (full density) plus a solid mean &plusmn; 1 SD band on top.

Real truth-value judgments by `scenario` (collapsed across `negation`/
`trigger`) — floor/ceiling-heavy and asymmetric in a way `rnorm()` samples
never quite are, which is exactly the case the "aura + SD-band" split is
meant to make legible.

```{r violin-sd-comparison, fig.height = 8}
#| fig.alt: >
#|   A two-by-two grid of violin plots of truth-value judgments, on a -2 to
#|   +2 scale, for the four scenarios true, false, undefined and critical.
#|   The top row shows full violins, the bottom row right-side half violins;
#|   the left column is the plain geom (geom_violin() and
#|   gghalves::geom_half_violin()), the right column this package's SD-band
#|   version. Every distribution is floor- and ceiling-heavy: mass piled at
#|   both ends of the scale with a narrow waist between. The plain violins
#|   render that as one flat silhouette, while the _sd versions overlay a
#|   solid, outlined mean plus-or-minus-one-SD band on a pale full-density
#|   aura, so how much of each scenario's spread falls inside one SD is
#|   readable directly.
demo_data <- d |> transmute(group = scenario, value = judgment)

violin_plain <- ggplot(demo_data, aes(x = group, y = value, fill = group)) +
  geom_violin() +
  labs(x = NULL, y = NULL) +
  guides(fill = "none")

violin_sd <- ggplot(demo_data, aes(x = group, y = value, fill = group)) +
  geom_violin_sd() +
  labs(x = NULL, y = NULL) +
  guides(fill = "none")

half_violin_plain <- ggplot(demo_data, aes(x = group, y = value, fill = group)) +
  geom_half_violin(side = "r") +
  labs(x = NULL, y = NULL) +
  guides(fill = "none")

half_violin_sd <- ggplot(demo_data, aes(x = group, y = value, fill = group)) +
  geom_half_violin_sd(side = "r") +
  labs(x = NULL, y = NULL) +
  guides(fill = "none")

(violin_plain + violin_sd) / (half_violin_plain + half_violin_sd)
```

### `style = "both"` / `"fill"` / `"outline"`

The SD band can be a solid fill, an outline, or both (default) — shared by
all three `_sd` geoms.

```{r violin-sd-style, fig.width = 12, fig.height = 5}
#| fig.alt: >
#|   The same four scenario violins drawn three times to compare the style
#|   argument. Left, style = both (the default): the SD band is a solid fill
#|   with an outline around it. Middle, style = fill: the same solid band
#|   with no outline. Right, style = outline: only the outlined band, with
#|   the pale full-density aura showing through where the fill used to be.
#|   The underlying distributions are identical across all three.
style_both <- ggplot(demo_data, aes(x = group, y = value, fill = group)) +
  geom_violin_sd(style = "both") +
  labs(x = NULL, y = NULL) +
  guides(fill = "none")

style_fill <- ggplot(demo_data, aes(x = group, y = value, fill = group)) +
  geom_violin_sd(style = "fill") +
  labs(x = NULL, y = NULL) +
  guides(fill = "none")

style_outline <- ggplot(demo_data, aes(x = group, y = value, fill = group)) +
  geom_violin_sd(style = "outline") +
  labs(x = NULL, y = NULL) +
  guides(fill = "none")

style_both + style_fill + style_outline
```

## `geom_split_violin_sd()`

Two `geom_half_violin_sd()` halves back-to-back, one per level of `split` —
no manual filtering or `side = "l"`/`"r"` calls needed.

```{r split-violin-data}
balanced_data <- d |> filter(trigger == "*again*")
split_mapping <- aes(x = scenario, y = judgment)
```

### Balanced

```{r split-violin-balanced, fig.height = 5}
#| fig.alt: >
#|   Back-to-back split violins of judgments, on a -2 to +2 scale, for the
#|   four scenarios. At each scenario the teal half for negation = without is
#|   drawn on the left and the crimson half for negation = with on the right.
#|   For true, the without half piles up at the top of the scale and the with
#|   half at the bottom; false reverses that; undefined puts both halves near
#|   the floor with a long tail upward; critical has without at the floor and
#|   with spread much higher.
ggplot(balanced_data, split_mapping) +
  geom_split_violin_sd(split_mapping, data = balanced_data, split = negation) +
  labs(x = NULL, y = "Judgment (rescaled -2/+2)")
```

### `flip = TRUE`

Same data, sides swapped.

```{r split-violin-flip, fig.height = 5}
#| fig.alt: >
#|   The same split violins as the previous figure with flip = TRUE, so each
#|   scenario's with half is now drawn on the left and its without half on
#|   the right. Every shape is mirrored across its scenario's centre line;
#|   the distributions themselves are unchanged, and the legend order swaps
#|   to match.
ggplot(balanced_data, split_mapping) +
  geom_split_violin_sd(split_mapping, data = balanced_data, split = negation, flip = TRUE) +
  labs(x = NULL, y = "Judgment (rescaled -2/+2)")
```

### Missing cell

Dropping `critical`/`with` entirely warns, not errors, and still renders a lone half-violin.

```{r split-violin-missing, fig.height = 5}
#| fig.alt: >
#|   The same split violins with the critical/with cell dropped from the
#|   data. The first three scenarios are unchanged. Critical now renders as a
#|   lone left-side half violin - the teal without half, its mass at the
#|   floor of the scale - with nothing drawn on the right, rather than the
#|   whole x position failing.
missing_data <- balanced_data |>
  filter(!(scenario == "critical" & negation == "with"))

ggplot(missing_data, split_mapping) +
  geom_split_violin_sd(split_mapping, data = missing_data, split = negation) +
  labs(x = NULL, y = "Judgment (rescaled -2/+2)")
```

### `scale = "count"` vs. `"area"`

`scale` only normalizes a side against itself, not against the other side —
`geom_split_violin_sd()`'s two sides are separate `Stat` computations, so
`scale` can only compare an x-level against *its own side's* other
x-levels, never against the other side's counts. The real cells above are
all ~100 trials, so the effect needs an example: subsampling `with` down
across `scenario` (holding `without` fixed) engineers the imbalance while
keeping every plotted judgment a real trial, rather than fabricating
ratings — `"count"` (left, default) lets the shrinking `with` N narrow that
side's violin; `"area"` (right) hides it entirely.

```{r split-violin-scale, fig.height = 5}
#| fig.alt: >
#|   Two panels of the same back-to-back split violins, with the with side
#|   subsampled to progressively smaller fractions across scenarios (all of
#|   true, then 70, 45 and 25 percent for false, undefined and critical).
#|   Left, scale = count (the default): each half's width is proportional to
#|   its own side's counts, so the shrinking with side narrows across the
#|   panel. Right, scale = area: every half is normalised to the same area,
#|   so the widths no longer carry that sample-size information at all.
set.seed(42)
with_fracs <- c(true = 1, false = .7, undefined = .45, critical = .25)
unbalanced_data <- bind_rows(
  balanced_data |> filter(negation == "without"),
  balanced_data |>
    filter(negation == "with") |>
    group_by(scenario) |>
    group_modify(~ slice_sample(.x, prop = with_fracs[[as.character(.y$scenario)]])) |>
    ungroup()
)

p_scale_count <- ggplot(unbalanced_data, split_mapping) +
  geom_split_violin_sd(split_mapping, data = unbalanced_data, split = negation) +
  labs(x = NULL, y = "Judgment (rescaled -2/+2)")

p_scale_area <- ggplot(unbalanced_data, split_mapping) +
  geom_split_violin_sd(
    split_mapping,
    data = unbalanced_data, split = negation, scale = "area"
  ) +
  labs(x = NULL, y = NULL) +
  guides(fill = "none")

p_scale_count + p_scale_area
```
