---
title: "theme_mt() and the SD-band violin geoms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{theme_mt() and the SD-band violin geoms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6
)
```

```{r setup}
library(emptyviz)
library(ggplot2)
library(gghalves)
library(patchwork)
use_theme_mt()
```

## `theme_mt()`

Extends `theme_minimal()` with markdown/HTML-aware text (via `ggtext`), a
bottom legend, and an automatic discrete color palette. Same data/labels
below, only the theme differs: `theme_gray()`, `theme_minimal()`, `theme_mt()`.

```{r theme-compare, fig.width = 13, fig.height = 6}
# theme_mt()'s discrete palette and geom_text() font default are both
# theme-level settings that resolve once, from whichever theme is active
# when the plot's scale/geom defaults are first constructed - not from
# whatever theme ends up attached via `+ theme_X()` afterward. Since this
# vignette's global default is theme_mt() (set by use_theme_mt() above),
# the gray/minimal panels need an explicit scale_fill_hue()/family = "sans"
# to force ggplot2's real defaults instead of silently inheriting
# theme_mt()'s.
theme_compare_data <- data.frame(
  condition = c("Baseline", "Weak Trigger", "Strong Trigger", "Control"),
  mean_rating = c(6.1, 4.8, 2.3, 6.5),
  se = c(0.15, 0.22, 0.28, 0.12)
)
theme_compare_data$condition <- factor(
  theme_compare_data$condition,
  levels = theme_compare_data$condition
)

build_theme_compare_plot <- function() {
  ggplot(theme_compare_data, aes(x = condition, y = mean_rating, fill = condition, color = condition)) +
    geom_col(alpha = .3) +
    geom_errorbar(aes(ymin = mean_rating - se, ymax = mean_rating + se), width = 0.2) +
    labs(
      title = "Acceptability ratings by condition",
      subtitle = "synthetic **data**, *n* = 45 per condition",
      x = "Condition", y = "Mean rating &plusmn;1 SE (1&ndash;7)",
      caption = "Test caption"
    ) +
    guides(fill = "none", color = "none")
}

value_label <- function(...) {
  geom_text(
    aes(label = sprintf("%.1f", mean_rating)),
    vjust = -1.8, color = "black", size = 5,
    ...
  )
}

p_theme_gray <- build_theme_compare_plot() + theme_gray() + scale_fill_hue() + value_label(family = "sans")
p_theme_minimal <- build_theme_compare_plot() + theme_minimal() + scale_fill_hue() + value_label(family = "sans")
p_theme_mt <- build_theme_compare_plot() + theme_mt(base_size = 12) + value_label()

p_theme_gray + p_theme_minimal + p_theme_mt
```

## `geom_violin_sd()` / `geom_half_violin_sd()`

Compound layers built on `geom_violin()`/`gghalves::geom_half_violin()`: a
low-alpha "aura" (full density) plus a solid mean &plusmn; 1 SD band on top.

```{r violin-sd-comparison, fig.height = 8}
set.seed(42)
demo_data <- rbind(
  data.frame(group = "A", value = rnorm(80, mean = 0, sd = 1)),
  data.frame(group = "B", value = c(rnorm(70, mean = 1.5, sd = 0.6), rnorm(10, mean = 5, sd = 0.4))), # bimodal-ish, with an outlier cluster
  data.frame(group = "C", value = rnorm(80, mean = -0.5, sd = 1.8))
)

violin_plain <- ggplot(demo_data, aes(x = group, y = value, fill = group)) +
  geom_violin(trim = FALSE) +
  labs(x = NULL, y = NULL) +
  guides(fill = "none")

violin_sd <- ggplot(demo_data, aes(x = group, y = value, fill = group)) +
  geom_violin_sd() +
  labs(x = NULL, y = NULL) +
  guides(fill = "none")

half_violin_plain <- ggplot(demo_data, aes(x = group, y = value, fill = group)) +
  geom_half_violin(side = "r", trim = FALSE) +
  labs(x = NULL, y = NULL) +
  guides(fill = "none")

half_violin_sd <- ggplot(demo_data, aes(x = group, y = value, fill = group)) +
  geom_half_violin_sd(side = "r") +
  labs(x = NULL, y = NULL) +
  guides(fill = "none")

(violin_plain + violin_sd) / (half_violin_plain + half_violin_sd)
```

### `style = "both"` / `"fill"` / `"outline"`

The SD band can be a solid fill, an outline, or both (default) — shared by
all three `_sd` geoms.

```{r violin-sd-style, fig.width = 12, fig.height = 5}
style_both <- ggplot(demo_data, aes(x = group, y = value, fill = group)) +
  geom_violin_sd(style = "both") +
  labs(x = NULL, y = NULL) +
  guides(fill = "none")

style_fill <- ggplot(demo_data, aes(x = group, y = value, fill = group)) +
  geom_violin_sd(style = "fill") +
  labs(x = NULL, y = NULL) +
  guides(fill = "none")

style_outline <- ggplot(demo_data, aes(x = group, y = value, fill = group)) +
  geom_violin_sd(style = "outline") +
  labs(x = NULL, y = NULL) +
  guides(fill = "none")

style_both + style_fill + style_outline
```

## `geom_split_violin_sd()`

Two `geom_half_violin_sd()` halves back-to-back, one per level of `split` —
no manual filtering or `side = "l"`/`"r"` calls needed. Synthetic data:
acceptability ratings by wh-extraction distance, split by grammaticality.

```{r split-violin-data}
make_ratings <- function(n, mean, sd) pmin(7, pmax(1, rnorm(n, mean, sd)))

set.seed(42)
balanced_data <- rbind(
  data.frame(distance = "Local", grammaticality = "grammatical", rating = make_ratings(45, 6.2, 0.8)),
  data.frame(distance = "Local", grammaticality = "ungrammatical", rating = make_ratings(45, 5.8, 0.9)),
  data.frame(distance = "Long-Distance", grammaticality = "grammatical", rating = make_ratings(45, 5.5, 1.0)),
  data.frame(distance = "Long-Distance", grammaticality = "ungrammatical", rating = make_ratings(45, 3.0, 1.1)),
  data.frame(distance = "Island", grammaticality = "grammatical", rating = make_ratings(45, 5.0, 1.0)),
  data.frame(distance = "Island", grammaticality = "ungrammatical", rating = make_ratings(45, 1.8, 0.9))
)
balanced_data$distance <- factor(
  balanced_data$distance,
  levels = c("Local", "Long-Distance", "Island")
)

# grammatical trial counts decline with distance (a plausible exclusion
# pattern - longer/harder items lose more trials to comprehension-check
# failures), while ungrammatical stays fixed. geom_split_violin_sd()'s two
# sides are separate Stat computations, so `scale` only normalizes a side's
# own x-levels against each other - NOT one side against the other. This
# data is built to show that (within-side, across x) effect honestly, not
# a cross-side one.
unbalanced_data <- rbind(
  data.frame(distance = "Local", grammaticality = "grammatical", rating = make_ratings(90, 6.2, 0.8)),
  data.frame(distance = "Local", grammaticality = "ungrammatical", rating = make_ratings(20, 5.8, 0.9)),
  data.frame(distance = "Long-Distance", grammaticality = "grammatical", rating = make_ratings(55, 5.5, 1.0)),
  data.frame(distance = "Long-Distance", grammaticality = "ungrammatical", rating = make_ratings(20, 3.0, 1.1)),
  data.frame(distance = "Island", grammaticality = "grammatical", rating = make_ratings(25, 5.0, 1.0)),
  data.frame(distance = "Island", grammaticality = "ungrammatical", rating = make_ratings(20, 1.8, 0.9))
)
unbalanced_data$distance <- factor(
  unbalanced_data$distance,
  levels = c("Local", "Long-Distance", "Island")
)

# Island ungrammatical trials dropped entirely - e.g. excluded post-hoc for
# a norming problem.
missing_data <- balanced_data[
  !(balanced_data$distance == "Island" & balanced_data$grammaticality == "ungrammatical"),
]

split_mapping <- aes(x = distance, y = rating)
```

### Balanced

```{r split-violin-balanced, fig.height = 5}
ggplot(balanced_data, split_mapping) +
  geom_split_violin_sd(split_mapping, data = balanced_data, split = grammaticality) +
  labs(x = NULL, y = "Acceptability (1-7)")
```

### `flip = TRUE`

Same data, sides swapped.

```{r split-violin-flip, fig.height = 5}
ggplot(balanced_data, split_mapping) +
  geom_split_violin_sd(split_mapping, data = balanced_data, split = grammaticality, flip = TRUE) +
  labs(x = NULL, y = "Acceptability (1-7)")
```

### Missing cell

Excluding Island's ungrammatical trials entirely warns, not errors, and
still renders a lone half-violin.

```{r split-violin-missing, fig.height = 5}
ggplot(missing_data, split_mapping) +
  geom_split_violin_sd(split_mapping, data = missing_data, split = grammaticality) +
  labs(x = NULL, y = "Acceptability (1-7)")
```

### `scale = "count"` vs. `"area"`

`scale` only normalizes a side against itself, not against the other side
(see code comment above) — `"count"` (left, default) lets declining
grammatical N narrow the violin; `"area"` (right) hides it.

```{r split-violin-scale, fig.height = 5}
p_scale_count <- ggplot(unbalanced_data, split_mapping) +
  geom_split_violin_sd(split_mapping, data = unbalanced_data, split = grammaticality) +
  labs(x = NULL, y = "Acceptability (1-7)")

p_scale_area <- ggplot(unbalanced_data, split_mapping) +
  geom_split_violin_sd(
    split_mapping,
    data = unbalanced_data, split = grammaticality, scale = "area"
  ) +
  labs(x = NULL, y = NULL) +
  guides(fill = "none")

p_scale_count + p_scale_area
```
